Binning Strategies and Related Loss for Large Data

### 2015-02-18 Paper.Rnw
first draft of subsection for 2d extension for rectangular binning

@@

We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values. This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density. This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins. The number of quantile based bins, $N_q ,= J$ , must be defined in order to assign values to bin centers. The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as 
%
\begin{eqnarray}\label{quantbin}
c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
\end{eqnarray}
%
\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.  To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus $\eta$ is the index for the ordered value where the next lower quantile based bin will be centered.

>>>>

The standard and random binning algorithms are easily extendible to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data tuples $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$. The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as 
%
\begin{eqnarray}\label{standrecbin}
b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
\end{eqnarray}
%
where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively. The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
%
\begin{eqnarray}\label{randrecbin}
b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
\end{eqnarray}
%
where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.

@@

We could also consider a empirical quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values. This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density. This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins. The number of quantile based bins, $N_q$ , must be defined in order to assign values to bin centers. The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as 
%
\begin{eqnarray}\label{quantbin}
c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
\end{eqnarray}
%
\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.  To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus $\eta$ is the index for the ordered value where the next lower quantile based bin will be centered.
